---
title: "Chapter 2 Graphics"
author: "Taqi"
#date: "4/18/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("../R/header.R")
library(future)
.src(1)
```

```{r, eval = FALSE, include = FALSE}
set.seed(23)
run <- F
if(run){
  save <- T
  if(save){
    ggsave(filename = "test.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```

```{r}
# Note: Add this feature to spectrum.scatterplot()!
# Creates a matrix title given the dimension and classification of the matrix (ex. 20x20 Unif(0,1))
.make_matrix_title <- function(N, class){
  #dim_str <- paste(N,"X",N, sep = " ")
  dim_str <- paste(N,"x",N, sep = "")
  title_str <- paste(dim_str, class, sep = " ")
  return(title_str)
}
# Create a shorter alias of this function
makeTitle <- .make_matrix_title 
# Example
makeTitle(N = 20, class = "Standard Normal")
```

# Section 1

## Spectrum Comparison

```{r}
set.seed(23)
run <- F
if(run)
{
  # Set parameters
  N <- 50
  # Generate random matrix and ensemble
  P <- RM_norm(N = N)
  ens <- RME_norm(N = N, size = 100)
  # Get the spectra plots
  P_plot <- P %>% spectrum.scatterplot(mat_str = makeTitle(N, class = "Standard Normal"))
  ens_plot <- ens %>% spectrum.scatterplot(mat_str = makeTitle(N, class = "Standard Normal"))
  # Glue the plots using patchwork
  comparison_plot <- P_plot + ens_plot
  comparison_plot
  # Save the plot
  save <- T
  if(save){
    ggsave(filename = "2-1-2_comparison.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```

## Order Scheme Comparison

```{r}
set.seed(23)
run <- F
if(run){
  # Some random matrices and ensembles
  ens <- RME_beta(N = 100, beta = 4, size = 5)
  # Specturm by order scheme
  spec_sign <- ens %>% spectrum_par(sort_norms = F)
  spec_norm <- ens %>% spectrum_par(sort_norms = T)
  # Generate the plots
  sign_plot <-
    spec_sign %>%
    order.scatterplot(component = Re) +
    labs(title = "Sign-ordered Spectrum")
  norm_plot <-
    spec_norm %>%
    order.scatterplot(component = Re) +
    labs(title = "Norm-ordered Spectrum")
  # Glue the plots using patchwork
  comparison_plot <- sign_plot + norm_plot
  # Save the plot
  save <- T
  if(save){
    ggsave(filename = "2-2-1_orderscheme.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```

# Section 2

# Order Statistics 1

```{r}
set.seed(23)
run <- T
if(run){
  # Set parameters
  N <- 10
  # Generate ensemble
  ens <- RME_norm(N = N, symm = T, size = 2000)
  if(F){  
    # Get spectrum 
    ens_spec <- ens %>% spectrum_par(order = 1:2)
    write.csv(ens_spec, "data/spec_order12.csv", row.names = F)
  } else{
    ens_spec <- read_csv("data/spec_order12.csv")
  }
  order_1 <- ens_spec %>% filter(Order == 1)
  order_2 <- ens_spec %>% filter(Order == 2)
  # Get plot
  ens_plot <- ens_spec %>% ggplot() +
    geom_histogram(data = order_1, mapping = aes(x = Norm, fill = "1"), bins = 50) +
    geom_histogram(data = order_2, mapping = aes(x = Norm, fill = "2"), bins = 50) +
    labs(fill = "Order", 
         title = paste("Distribution of the Largest Eigenvalues of a", 
                       .makeTitle(N, "Symmetric Normal Matrix")))
  ens_plot
  # Save plot
  save <- T
  if(save){
    ggsave(filename = "2-2-1_spec_order12.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```

## Order Statistics 2

```{r}
set.seed(23)
run <- T
if(run){
  # Set parameters
  N <- 50
  # Generate ensemble
  ens <- RME_norm(N = N, symm = T, size = 300)
  if(T){  
    # Get spectrum 
    ens_spec <- ens %>% spectrum_par()
    write.csv(ens_spec, "data/spec_order_density.csv", row.names = F)
  } else{
    ens_spec <- read_csv("data/spec_order_density.csv")
  }
  # Get plot
  ens_plot <- ens_spec %>%
    ggplot(mapping = aes(x = Norm, group = Order, color = Order)) +
    geom_density(kernel = "gaussian") +
    scale_color_gradient(low = "red", high = "blue") +
      labs(fill = "Order", 
           title = paste("Distribution of the Eigenvalues of a", 
                         .makeTitle(N, "Symmetric Normal Matrix by Order")))
  ens_plot
  # Save plot
  save <- T
  if(save){
    ggsave(filename = "2-2-1_spec_order_density.pdf", path = "chap2", width = 10.5, height = 7)
  }
}

```

## Order Statistics 3

```{r}
set.seed(23)
run <- T
if(run){
  # Set parameters
  N <- 50
  # Generate ensemble
  ens <- RME_norm(N = N, symm = T, size = 300)
  if(F){  
    # Get spectrum 
    ens_spec <- ens %>% spectrum_par()
    write.csv(ens_spec, "data/spec_order_density.csv", row.names = F)
  } else{
    ens_spec <- read_csv("data/spec_order_density.csv")
  }
  # Get plot
  ens_plot <- order.summary(ens_spec, Variance_Re)
  ens_plot
  # Save plot
  save <- T
  if(save){
    ggsave(filename = "2-2-1_spec_order_summary.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```




# Section 3

## Beta: Symmetric Spectrum

```{r}
set.seed(23)
run <- F
if(run){
  # Set parameters
  N <- 50
  # Generate a spectrum of a symmetric matrix ensemble
  ens_plot <- RME_beta(N = N, beta = 4, size = 100) %>% 
    spectrum.scatterplot(mat_str = makeTitle(N, "Beta-4"))
  # Save the plot
  save <- T
  if(save){
    ggsave(filename = "2-3-1_symmetric_spectrum.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```

## Semicircle Distribution

```{r}
set.seed(23)
run <- T
if(run){
  # Set parameters
  N <- 50
  # Get the beta ensemble
  ens <- RME_beta(N = N, beta = 2, size = 1000)
  # Plot distribution of eigenvalues
  ens_spec <- ens %>% spectrum(components = T) 
  ens_plot <- ens_spec %>% spectrum.histogram(component = "Re", bins = 100)
  # Save the plot
  save <- T
  if(save){
    ggsave(filename = "2-3-2_semicircle.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```

## Section 4

## Uniform Ensemble Spectrum

```{r}
set.seed(23)
run <- F
if(run){
  # Set parameters
  N <- 50
  # Obtain the uniform(0,1) spectrum plot
  ens_plot <- RME_unif(N = N, min = 0, max = 1, size = 1000) %>% 
    spectrum.scatterplot(mat_str = makeTitle(N, class = "Unif(0,1)"))
  # Save the plot
  save <- T
  if(save){
    ggsave(filename = "2-4_unif01_spec.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```

## Normal Ensemble Spectrum

```{r}
set.seed(23)
run <- F
if(run){
  # Set parameters
  N <- 50
  # Obtain the standard normal spectrum plot
  ens_plot <- RME_norm(N = N, size = 500) %>% 
    spectrum.scatterplot(mat_str = makeTitle(N, class = "Standard Normal"))
  ens_plot
  # Save the plot
  save <- T
  if(save){
    ggsave(filename = "2-4_normal_spec.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```

## Normal Complex Hermitian Ensemble Spectrum

```{r}
set.seed(23)
run <- T
if(run){
  # Set parameters
  N <- 200
  # Obtain the complex hermitian standard normal spectrum plot
  ens_plot <- RME_norm(N = N, cplx = T, herm = T, size = 500) %>% 
    spectrum.scatterplot(mat_str = makeTitle(N, class = "Complex Hermitian Standard Normal"))
  ens_plot
  # Save the plot
  save <- T
  if(save){
    ggsave(filename = "2-4_cplxherm_normal_spec.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```

## Stochastic Ensemble Spectrum

```{r}
set.seed(23)
run <- F
if(run){
  # Set parameters
  N <- 50
  # Obtain the stochastic spectrum plot
  ens_plot <- RME_stoch(N = N, size = 50) %>% 
    spectrum.scatterplot(mat_str = makeTitle(N, class = "Stochastic"))
  ens_plot
  # Save the plot
  save <- T
  if(save){
    ggsave(filename = "2-4_stoch_spec.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```

## Symmetric Stochastic Ensemble Spectrum

```{r}
set.seed(23)
run <- F
if(run){
  # Set parameters
  N <- 50
  # Obtain the symmetric stochastic spectrum plot
  ens_plot <- RME_stoch(N = N, symm = T, size = 50) %>% 
    spectrum.scatterplot(mat_str = makeTitle(N, class = "Symmetric Stochastic"))
  ens_plot
  # Save the plot
  save <- T
  if(save){
    ggsave(filename = "2-4_symmstoch_spec.pdf", path = "chap2", width = 10.5, height = 7)
  }
}
```

## Erdos-Renyi 2nd Largest Eigenvalue

```{r}
# ===================================================
#                       Setup
# ===================================================
# Labels a generic array with a value column
.label_array <- function(array, value, label, size){
  label_column <- data.frame(x = rep(value, size))
  colnames(label_column) <- label
  cbind(array, label_column)
}
# Creates a tidy, labelled p-Erdos ensemble spectrum
erdos_spectrum <- function(p, N, size){
  ensemble_spectrum <- RME_erdos(N = N, p = p, size = size) %>% spectrum_par(order = 2:3)
  ensemble_spectrum %>% .label_array(value = p, label = "p", size = nrow(ensemble_spectrum))
}
# ===================================================
#                     Simulation
# ===================================================
set.seed(23)
run <- T
if(run){
# Global parameters
N <- 50
size <- 2
#size <- 50
ps <- seq(0.005, 0.9995, 0.005)
# ===================================================
# Generate the p-ensembles
erdos_ensemble <- purrr::map_dfr(ps, erdos_spectrum, N, size)
# ===================================================
# Plot
erdos_lam2 <-
  erdos_ensemble %>%
  ggplot(mapping = aes(x = Re, y = Im, color = p)) +
  geom_point() +
  scale_color_viridis_c() +
  labs(title = "2nd Largest Eigenvalue of p-Erdos Ensembles by p")
save <- T
if(save){ggsave(filename = "2-4_erdos_lam2.pdf", path = "chap2", width = 9, height = 7)}
}
```
