%% TeX Preamble
\documentclass[12pt,twoside]{reedthesis}
\input{preamble/preamble.tex} % Input supporting TeX header files
\newcommand{\achknowledge}[1]{\noindent \textit{#1},}

%% Thesis Preamble
\title{Spectral Statistics of Random Matrices}
\author{Ali Taqi}
\date{May 2021}
\division{Mathematical and Natural Sciences}
\advisor{Jonathan M. Wells}
\department{Mathematics}

% TeX Settings
\setlength{\parskip}{0pt}
\setlength{\headheight}{15pt}

% The fun begins
\begin{document}
  \maketitle
  \frontmatter % this stuff will be roman-numbered
  \pagestyle{empty} % this removes page numbers from the frontmatter

     \chapter*{Acknowledgements}

	\achknowledge{Nate} thank you for being an amazing mentor, and for being kind and insightful every step of the way.
	Also, thank you for introducing me to the wonderful worlds of probability theory and random matrices. \\

	\achknowledge{Kelly} taking introductory statistics with you was such an instrumental experience in my journey as a statistician.
  Thank you for teaching me $R$ and being my wonderful (re)introduction to the world of statistics.
	I owe all these awesome visualizations to you. \\

	\achknowledge{Ang\'elica} I really appreciate your encouragement and the values of persistence you tried to instill in me.
  In my darkest days, I always tried to fall back on and recall what you told me.
  By and large, I really do believe that was the reason I kept going. \\

	\achknowledge{Samuel} thank you for bringing the love and joy I needed to get through this year.
  No matter what or when, I know I can always count on our memories to fill me with glee and elation.
	I just know you are going to do amazing things and I cannot wait to hear all about it. \\

  \achknowledge{Lucy} I cannot explain the odds of the circumstances in which our paths crossed.
  I guess sometimes we just have to marvel at the serendipity life brings us.
  You will always be my truest friend. \\

  \achknowledge{Candice} I do not think I can overstate the level with which you shaped me both academically and personally.
  I still remember the day you picked me from the Portland airport.
	Though I had just arrived from a 7000 mile journey, you made it such that it felt like I was already home.  \\

  \achknowledge{Juce} I will always treasure our countless memories of joy, love, and friendship.
   No friendship have I ever had was as genuine and consequential as ours.
   I cannot wait to see what we both grow to become.  \\

 	\noindent \textit{And to all my friends who know themselves} -- thank you for all your love and support!

%     \chapter*{Preface}
% 	This is an example of a thesis setup to use the reed thesis document class.

    \chapter*{List of Abbreviations}

	\begin{table}[h]
	\centering % You could remove this to move table to the left
	\begin{tabular}{ll}
		\textbf{r.v}  	  &  Random Variable \\
		\textbf{i.i.d}  	&  Independent and Identically Distributed \\
		%\textbf{h.e.d}    &  Homogenously and Explicitly Distributed \\
		\textbf{pdf}      &  Probability Distribution Function \\
		\textbf{cdf}      &  Cumulative Distribution Function \\
		%\textbf{w.r.t}		&  with respect to
		\end{tabular}
	\end{table}

    \tableofcontents

    \listoftables \hfill

    \begin{center}

    \Ddisttable \newline

    \hfill \vspace{2em}

    \spectrumschemetable \newline

    \hfill \vspace{2em}

    \dispersiontable \newline

    \hfill \vspace{2em}

    \pairingschemetable \newline

    \end{center}

%    \listoffigures

% If your abstract is longer than a page, there may be a formatting issue.
    \chapter*{Abstract}
	On their own, random variables exhibit deterministic properties regarding their uncertainty.
	The same generalization applies to random matrices, which are matrices where some or all of its entries are random variables.
	One particular random matrix statistic worth investigating is the distribution of its eigenvalues or its spectrum.
	In this thesis, we explore various classes of random matrices and their corresponding spectral statistics. Namely, we observe their spectra and dispersions.
	That will be demonstrated with simulations, in which we survey several results in random matrix theory and related fields.

	\chapter*{Dedication}
  \vspace{15em}
  \begin{center}
    For my mother \textit{Eman}, who I am eternally indebted to.
  \end{center}

  \mainmatter % here the regular arabic numbering starts
  \pagestyle{fancyplain} % turns page numbering back on

\input{chapters/introduction.tex}

% Double spacing: if you want to double space, or one and a half space, uncomment one of the following lines.
% You can go back to single spacing with the \singlespacing command.
% \onehalfspacing
% \doublespacing

% Beginning of Chapters
\input{chapters/chapter1.tex}
\input{chapters/chapter2.tex}
\input{chapters/chapter3.tex}
\input{chapters/chapter4.tex}
% End of Chapters

\input{chapters/conclusion.tex}

% Appendix
    \appendix
      \input{appendix/math_appendix.tex}
      \input{appendix/algorithms_appendix.tex}
      \input{appendix/code_appendix.tex}
      \input{appendix/ratios.tex}

%This is where endnotes are supposed to go, if you have them.
%I have no idea how endnotes work with LaTeX.

  \backmatter % backmatter makes the index and bibliography appear properly in the t.o.c...

% if you're using bibtex, the next line forces every entry in the bibtex file to be included
% in your bibliography, regardless of whether or not you've cited it in the thesis.
    \nocite{*}

% Rename my bibliography to be called "Works Cited" and not "References" or ``Bibliography''
% \renewcommand{\bibname}{Works Cited}

%\bibliographystyle{bsts/mla-good} % there are a variety of styles available;
%\bibliographystyle{plainnat}
% replace ``plainnat'' with the style of choice. You can refer to files in the bsts or APA
% subfolder, e.g.
 \bibliographystyle{APA/apa-good}  % or
 \bibliography{thesis}
 % Comment the above two lines and uncomment the next line to use biblatex-chicago.
 %\printbibliography[heading=bibintoc]

% Finally, an index would go here... but it is also optional.
\end{document}
