---
title: "Matrix Refactor"
author: "Ali Taqi"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", 
                      message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(patchwork)
# Load files
source(file = "R/matrices_new.R")
```

```{r}
# Global parameters
M <- 3
```

## Normal Matrices

```{r}
# Set seed
set.seed(23)
# Set parameters
mu <- 1
sd <- 2
# Generate matrixx
P <- RM_normal(M, normal_args = c(mu, sd), symm = F)
P
```

### Symmetric Normal Matrices


```{r}
# Set seed
set.seed(23)
# Set parameters
mu <- 1
sd <- 2
# Generate matrixx
P <- RM_normal(M, normal_args = c(mu, sd), symm = T)
P
```

\newpage

## Stochastic Matrices

### Sparse

```{r}
# Set seed
set.seed(23)
# Set parameters
mu <- 1
sd <- 2
# Generate matrixx
P <- RM_stoch(M, symm = F, sparsity = T)
P
```

### Non-sparse

```{r}
# Set seed
set.seed(23)
# Set parameters
mu <- 1
sd <- 2
# Generate matrixx
P <- RM_stoch(M, symm = F, sparsity = F)
P
```

\newpage

### Symmetric Stochastic Matrices

```{r}
# Set seed
set.seed(23)
# Set parameters
mu <- 1
sd <- 2
# Generate matrixx
P <- RM_stoch(M, symm = T, sparsity = T)
P
```

### Non-sparse

```{r}
# Set seed
set.seed(23)
# Set parameters
mu <- 1
sd <- 2
# Generate matrixx
P <- RM_stoch(M, symm = T, sparsity = F)
P
```

\newpage

## Tridiagonal Matrices

```{r}
# Set seed
set.seed(23)
# Generate matrix
P <- RM_trid(M)
P
```

## p-Sparse Matrices

```{r}
# Set seed
set.seed(23)
# Set parameters
p <- 0.2
# Generate matrix
P <- RM_erdos(M, p, stoch = F)
```

### Stochastic p-Sparse Matrices

```{r}
# Set seed
set.seed(23)
# Set parameters
p <- 0.2
# Generate matrix
P <- RM_erdos(M, p, stoch = T)
```

\newpage

Suppose we have a $M \times M$ square matrix $\bf P$ (for some $M \in \mathbb{N}$) on a field $F$. We notate $\mathbf{P} \in \mathcal{M}_F[M^2]$.

Take $\mathbf{P} \in \mathcal{M}_F[M^2]$.

## Structural Properties of Matrices

If $\bf P$ is symmetric, then its upper triangle is equal to the lower triangle.

If $\bf P$ is tridiagonal, then it is a band matrix of width 1.

\bigskip

## Entry-wise Properties of Matrices

If $\bf P$ is row-stochastic, then $\forall i : \sum_j p_{ij} = 1$.

`RM_stoch <- function(M, symm = F, sparsity = F){...}`

\medskip

If $\bf P$ is $\mathcal{N}(\mu,\sigma^2)$, then its entries satisfy $p_{ij} \sim \mathcal{N}(\mu,\sigma^2)$.

`RM_normal <- function(M, normal_args = c(0,1), symm = F){...}`

\medskip

If $\bf P$ is $p-\text{sparse}$, then $\forall i, j \in S_M : p_{ij}/c \sim \text{Bern}(p)$ for some $c \in \mathbb{R}$.

`RM_erdos <- function(M, p_sparse){...}` 