---
title: "Eigenvectors of Symmetric Matrices"
author: "Ali Taqi"
date: "11/4/2020"
output: pdf_document
header-includes:
  - \newcommand{\R}{\mathbb{R}}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", 
                      message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(patchwork)
library(ggfortify)
library(matrixcalc)
library(zeallot)
# Load files
source(file = "R/sym_eigen.R")
source(file = "R/eigenmetrics.R")
source(file = "R/eigenvectors.R")
```

# Computational Evidence: Real Symmetric Matrices have Real Eigenvectors

In this document, we hope to show that given any arbitrary element of the set of $M \times M$ symmetric matrices, denote it $S \in \mathcal{SM}_\R[M \times M]$ has a set of real eigenvectors $[\lambda_i] \in \R^M$.

To simulate a generic element $S$, we use the following method:

(1) First, pick some $f \in [0,1]$, letting it denote the fraction of positive entries of $S$. That is;
$$\text{Want: }\, f \approx \frac{|{s_{ij} > 0}|}{M^2} $$
We hope to show that our condition is invariant to the value of $f$, since there is the possibility that the sign proportions of our matrix $S$ influences the $\det (S)$. 

(2) To simulate a symmetric matrix $S$ with a fraction of positive entries $f$, we will sample from the distribution:
$$s_{ij} \sim \text{Unif}(f-1,f)$$

(3) To not constrict the sizes of $|s_{ij}|$, we will add an $\epsilon$ term and scale our endpoints to preserve the fraction $f$.
$$s_{ij} \sim \text{Unif}(\epsilon(f-1),\,\epsilon f)$$

```{r}
unif_fpos <- function(M,f,ep){
  # unless specifically initialized, a random fraction will be chosen
  if(F){
    f <- runif(1,0,1)
    paste("f: ",f,sep="")
  }
  b <- f
  a <- (f-1)
  dist <- data.frame(x = runif(M**2, ep*a, ep*b))
  dist <- dist %>% mutate(x_neg = ifelse(x < 0,yes = 1, no = 0))
  dist
}
```


(4) Having our uniform distribution, we will generate $M^2$ entries and insert them in the matrix $S$. Then, we delete the lower triangular matrix, then duplicate the entries from the upper triangle to the lower triangle. 

```{r}
make_symm <- function(dist){
  N <- sqrt(length(dist$x))
  P <- matrix(data = dist$x, nrow = N, ncol = N)
  LT <- lower.tri(P)
  UT <- upper.tri(P)
  P[LT] <- P[UT]
  P
}
```

(5) Now, if we let $f \sim \text{Unif}(0,1)$ and let $\epsilon \to \infty$, we can well approximate $S \in \mathcal{SM}_\R[M \times M]$.

\newpage

# Simulation

```{r, cache=T}
# add matrix so that we plot(ep,M) on xy plane and color value of diff to show 2d relation of convergence of our method
f <- 0.5

M_vec <- sample(2:200, 100,replace = F)
ep_vec <- sample(2:200, 100,replace = F)
table <- data.frame(M = M_vec, ep = rep(ep_vec,length(M_vec)))

check_real_eigenvectors <- function(M,ep,f){
  S <- RM_symm(M,f,ep)
  S_real <- eigenvectors_real(eigen_frame(S)) # use function for eigenvectors.R
  S_real
}

eigenvecs_real <- rep(NA, length(table$M))

for(i in 1:length(table$M)){
  eigenvecs_real[i] <- check_real_eigenvectors(table[i,][1],table[i,][2],f)
}
table <- cbind(table,eigenvecs_real)
```

```{r}
head(table)
```

```{r}
ggplot() + 
  geom_point(data = table, aes(x=M,y=ep, color = factor(eigenvecs_real))) +
  labs(color = "EV Real", title = "f = 0.5")
```


```{r, cache=T}
# add matrix so that we plot(ep,M) on xy plane and color value of diff to show 2d relation of convergence of our method
f <- 0.1

M_vec <- sample(2:200, 100,replace = F)
ep_vec <- sample(2:200, 100,replace = F)
table <- data.frame(M = M_vec, ep = rep(ep_vec,length(M_vec)))

check_real_eigenvectors <- function(M,ep,f){
  S <- RM_symm(M,f,ep)
  S_real <- eigenvectors_real(eigen_frame(S)) # use function for eigenvectors.R
  S_real
}

eigenvecs_real <- rep(NA, length(table$M))

for(i in 1:length(table$M)){
  eigenvecs_real[i] <- check_real_eigenvectors(table[i,][1],table[i,][2],f)
}
table <- cbind(table,eigenvecs_real)
```

```{r}
ggplot() + 
  geom_point(data = table, aes(x=M,y=ep, color = factor(eigenvecs_real))) +
  labs(color = "EV Real", title = "f = 0.1")
```

```{r, cache=T}
# add matrix so that we plot(ep,M) on xy plane and color value of diff to show 2d relation of convergence of our method
f <- 0.9

M_vec <- sample(2:200, 100,replace = F)
ep_vec <- sample(2:200, 100,replace = F)
table <- data.frame(M = M_vec, ep = rep(ep_vec,length(M_vec)))

check_real_eigenvectors <- function(M,ep,f){
  S <- RM_symm(M,f,ep)
  S_real <- eigenvectors_real(eigen_frame(S)) # use function for eigenvectors.R
  S_real
}

eigenvecs_real <- rep(NA, length(table$M))

for(i in 1:length(table$M)){
  eigenvecs_real[i] <- check_real_eigenvectors(table[i,][1],table[i,][2],f)
}
table <- cbind(table,eigenvecs_real)
```

```{r}
ggplot() + 
  geom_point(data = table, aes(x=M,y=ep, color = factor(eigenvecs_real))) +
  labs(color = "EV Real", title = "f = 0.9")
```

