---
title: "Refactoring matrices.R"
author: "Ali Taqi"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", 
                      message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(patchwork)
# Load files
source(file = "../R/matrices.R")
```

```{r, include = F}
bloud <- T
bplot <- F
```


## Normal Matrices

### Basic Example

```{r}
# Set seed
set.seed(23)
# Set parameters
M <- 5
mu <- 0
sd <- 2
normal_args <- c(mu, sd)
# Generate matrix
P <- RM_normal(M, normal_args = c(mu, sd), symm = T)
```

```{r}
entries_P <- vectorize_matrix(P)
normal_params(entries_P)
```

```{r, echo = F}
if(bloud){P}
is_symmetric(P)
if(bplot){visualize_normal_entries(P, normal_args)}
```

### Symmetric Normal Matrices

```{r, echo = F}
RM_normal <- function(M, normal_args = c(0,1), symm = F){
  # Extract parameters
  mu <- normal_args[1]
  sd <- normal_args[2]
  # Create [M x M] transition matrix
  P <- matrix(rep(NA, M * M), ncol = M)  
  # Generate rows
  for(i in 1:M){
    P[i,] <- rnorm(n = M, mean = mu, sd = sd)
  }
  # Make symmetric if prompted
  if(symm == T){P <- equalize_triangles(P)}
  # Return the matrix
  P
}
```

```{r}
# Set seed
set.seed(23)
# Set parameters
M <- 5
mu <- 0
sd <- 1
normal_args <- c(mu, sd)
# Generate matrix
P <- RM_normal(M, normal_args, symm = T)
```

```{r, echo = T, message = T}
if(bloud){P}
is_symmetric(P)
if(!bplot){visualize_normal_entries(P, normal_args)}
```

\newpage

## Stochastic Matrices

### Sparse Stochastic Matrices

```{r}
# Set seed
set.seed(23)
# Set parameters
M <- 3
# Generate matrix
P <- RM_stoch(M, symm = F, sparsity = T)
```

```{r, echo = F}
if(bloud){P}
is_row_stochastic(P)
is_symmetric(P)
```



### Non-sparse Stochastic Matrices

```{r}
# Set seed
set.seed(23)
# Set parameters
M <- 3
# Generate matrix
P <- RM_stoch(M, symm = F, sparsity = F)
```

```{r, echo = F}
if(bloud){P}
is_row_stochastic(P)
is_symmetric(P)
```

\newpage

### Symmetric Stochastic Matrices

### Sparse Symmetric Stochastic Matrices

```{r}
# Generate random stochastic matrix of size M, with choice of row function {r_stochastic, r_zeros}
RM_stoch <- function(M, symm = F, sparsity = F, tune_D = T){
  P <- matrix(rep(NA, M * M), ncol = M)  # create [M x M] transition matrix
  if(sparsity){row_fn <- r_zeros} else {row_fn <- r_stochastic} # choose row function
  # Generate rows
  for(i in 1:M){P[i,] <- row_fn(M)}
  # Make symmetric (if prompted)
  if(symm == T){
    # Add Transpose
    P <- (P + t(P))/2
    # Set diagonal such that rows sum to 1
    diag <- rep(0, ncol(P))
    for(i in 1:nrow(P)){
      row <- P[i, ]
      diag[i] <- (1 - sum(nondiagonal_entries(row, i)))
    }
    diag(P) <- diag
    # Retune the diagonals in case any were negative
    if(tune_D){P <- retune_diagonal(P)}
  }
  # Return the matrix
  P
}
```

```{r}
# Set seed
set.seed(23)
# Set parameters
M <- 7
# Generate matrix
P <- RM_stoch(M, symm = T, sparsity = T)
if(bloud){P}
is_row_stochastic(P)
row_sums(P)
is_symmetric(P)
```

```{r}
P
P_ <- spread_negative_diagonal(P, 7)
P_
row_sums(P_)
col_sums(P_)
```


### Non-sparse Symmetric Stochastic Matrices

```{r}
# Set seed
set.seed(23)
# Set parameters
M <- 3
mu <- 1
sd <- 2
# Generate matrix
P <- RM_stoch(M, symm = T, sparsity = F)
```

```{r, echo = F}
if(bloud){P}
is_row_stochastic(P)
is_symmetric(P)
```

\newpage

## Tridiagonal Matrices

### Basic example

```{r}
# Set seed
set.seed(23)
# Set parameters
M <- 3
# Generate matrix
# Need not be symmetric : fix laater
P <- RM_trid(M)
```

```{r, echo = F}
if(bloud){P}

```


### Symmetric Tridiagonal Matrices

```{r}
# Set seed
set.seed(23)
# Set parameters
M <- 3
# Generate matrix
P <- RM_trid(M)
# Need not be symmetric : fix later
```

## p-Sparse Matrices

### Basic example

```{r}
# Set seed
set.seed(23)
# Set parameters
M <- 3
p <- 0.2
# Generate matrix
P <- RM_erdos(M, p, stoch = F)
```

```{r, echo = F}
if(bloud){P}
is_row_stochastic(P)
```

### Stochastic p-Sparse Matrices

```{r}
# Set seed
set.seed(23)
# Set parameters
M <- 3
p <- 0.2
# Generate matrix
P <- RM_erdos(M, p, stoch = T)
```

```{r, echo = F}
if(bloud){P}
is_row_stochastic(P)
```

## Old Strategy

```{r}
# Generate random stochastic matrix of size M, with choice of row function {r_stochastic, r_zeros}
RM_stoch <- function(M, symm = F, sparsity = F, tune_D = T){
  P <- matrix(rep(NA, M * M), ncol = M)  # create [M x M] transition matrix
  if(sparsity){row_fn <- r_zeros} else {row_fn <- r_stochastic} # choose row function
  # Generate rows
  for(i in 1:M){P[i,] <- row_fn(M)}
  # Make symmetric (if prompted)
  if(symm == T){
    # Add Transpose
    P <- (P + t(P))/2
    # Set diagonal such that rows sum to 1
    diag <- rep(0, ncol(P))
    for(i in 1:nrow(P)){
      row <- P[i, ]
      diag[i] <- (1 - sum(nondiagonal_entries(row, i)))
    }
    diag(P) <- diag
    # Retune the diagonals in case any were negative
    if(tune_D){P <- retune_diagonal(P)}
  }
  # Return the matrix
  P
}
```

```{r}
# Check for negative diagonal entries then repair if needed
retune_diagonal <- function(P){
  for(i in 1:nrow(P)){
    if(P[i,i] < 0){
      P <- spread_negative_diagonal(P, row_idx = i)
    }
  }
  P
}
# If a diagonal entry at the i^th vector (row/column) has a negative value, renormalize the row and column of that vector index
spread_negative_diagonal <- function(P, row_idx){
  i <- row_idx
  # Get troubled row
  row_i <- P[i, ]
  # Find positive entries
  POS <- which(row_i > 0)
  # Spread out the negative diagonal entry across each positive entry
  diag_entry <- P[i,i]
  portion <- diag_entry/length(POS)
  # Renormalize the row by adding portions to positive entries
  P[i, POS] <- P[i, POS] - portion
  renormalized <- P[i, ]
  # Set column and row to renormalized, nonnegative stochastic row/column pair
  P[i, ] <- renormalized
  P[,i] <- t(renormalized)
  # Set diagonal to 0
  P[i,i] <- 0
  # Return matrix
  P
}
```

```{r}
# If a diagonal entry at the i^th vector (row/column) has a negative value, renormalize the row and column of that vector index
spread_negative_diagonal <- function(P, row_idx){
  i <- row_idx
  # Get troubled row
  row_i <- P[i, ]
  # Find positive entries
  POS <- which(row_i > 0)
  # Spread out the negative diagonal entry across each positive entry
  diag_entry <- P[i,i]
  portion <- diag_entry/length(POS)
  # Renormalize the row by adding portions to positive entries
  P[i, POS] <- P[i, POS] <- portion
  # Set diagonal to 0 
  P[i,i] <- 0
  renormalized <- P[i, ]/sum(P[i, ])
  # Set column and row to renormalized, nonnegative stochastic row/column pair
  P[i, ] <- renormalized
  P[,i] <- t(renormalized)
  # Set diagonal to 0
  # Return matrix
  P
}
```

\newpage
\newpage

# Notation

Suppose we have a $M \times M$ square matrix $\bf P$ (for some $M \in \mathbb{N}$) on a field $F$. We notate $\mathbf{P} \in \mathcal{M}_F[M^2]$.

Take $\mathbf{P} \in \mathcal{M}_F[M^2]$.

## Structural Properties of Matrices

If $\bf P$ is symmetric, then its upper triangle is equal to the lower triangle.

If $\bf P$ is tridiagonal, then it is a band matrix of width 1.

\bigskip

## Entry-wise Properties of Matrices

If $\bf P$ is row-stochastic, then $\forall i : \sum_j p_{ij} = 1$.

`RM_stoch <- function(M, symm = F, sparsity = F){...}`

\medskip

If $\bf P$ is $\mathcal{N}(\mu,\sigma^2)$, then its entries satisfy $p_{ij} \sim \mathcal{N}(\mu,\sigma^2)$.

`RM_normal <- function(M, normal_args = c(0,1), symm = F){...}`

\medskip

If $\bf P$ is $p-\text{sparse}$, then $\forall i, j \in S_M : p_{ij}/c \sim \text{Bern}(p)$ for some $c \in \mathbb{R}$.

`RM_erdos <- function(M, p_sparse){...}` 