---
title: "Order Statistics Simulations"
author: "Taqi"
date: "3/9/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source("../../R/header.R")
.src(2)
```

## Order Statistics Simulations

```{r}
# Simulating function
sim_orders <- function(N, iter, sample_fn, ...){
  order_stats <- function(N, ...){
    x <- sample_fn(N, ...)
    data.frame(value = sort(x), order = (1:length(x))) # Get order statistics
  }
  map_dfr(rep(N, iter), order_stats, ...)
}
# Simulating function for complex versions
sim_orders_cplx <- function(N, iter, sample_fn, ...){
  order_stats <- function(N, ...){
    x <- abs(sample_fn(N, ...) + 1i * sample_fn(N, ...)) # Norm of complex version
    data.frame(value = sort(x), order = (1:length(x))) # Get order statistics
  }
  map_dfr(rep(N, iter), order_stats, ...)
}
# Plot a dataframe of order data statistics
orders_plot <- function(orders, N){
  # Plotting parameters
  #order_cols <- scales::seq_gradient_pal("blue", "red", "Lab")(seq(0, 1, length.out = N))
  # Plot
  orders %>%
    ggplot(aes(value, order, color = order)) + 
    geom_point(show.legend = FALSE) +
    #scale_y_continuous(breaks = 1:N) +
    scale_color_gradient(low = "deepskyblue3", high = "firebrick3")
}
```

```{r}
# Simulation parameters
iter <- 500
N <- 100
```


## Uniform Distribution Order Statistics

```{r}
# Simulate uniformly distributed order statistics
sim_orders(N, iter, sample_fn = runif, min = -1, max = 1) %>% orders_plot(N)
```

### Norm Version

```{r}
# Define norm of random uniform distribution function
NORM_runif <- function(n, min, max){abs(runif(n, min, max))}
# Simulate uniformly distributed order statistics
sim_orders(N, iter, sample_fn = NORM_runif, min = -1, max = 1) %>% orders_plot(N)
```

## Normal Distribution Order Statistics

```{r}
# Simulate normally distributed order statistics
sim_orders(N, iter, sample_fn = rnorm) %>% orders_plot(N)
```

### Norm Version

```{r}
# Define norm of random uniform distribution function
NORM_rnorm <- function(n, ...){abs(rnorm(n, ...))}
# Simulate normally distributed order statistics
sim_orders(N, iter, sample_fn = NORM_rnorm) %>% orders_plot(N)
```


## Norm of Uniformly Distributed Complex Number Order Statistics

```{r}
# Simulate
sim_orders_cplx(N, iter, runif, min = -1, max = 1) %>% orders_plot(N)
```

## Norm of Normally Distributed Complex Number Order Statistics

```{r}
# Simulate
sim_orders_cplx(N, iter, rnorm, mean = 0, sd = 1) %>% orders_plot(N)
```

```{r}
# Source code for simulating from the semicircle distribution
source("semicircle.R")
```

## The Semicircle Distribution

```{r}
sample <- data.frame(value = sample_semicircle(n = 500000, R = 1))
sample %>%
  ggplot() +
  geom_histogram(aes(x = value), bins = 300)
```

## Semicircle Distribution Order Statistics

```{r}
# Simulate
sim_orders(N, iter, sample_semicircle, R = 1) %>% orders_plot(N)
```

## Norm of Semi-Circle Distributed Complex Number Order Statistics

```{r}
# Simulate
sim_orders_cplx(N, iter, sample_semicircle, R = 1) %>% orders_plot(N)
```

