---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```

# Matrices

Random matrices can either be explicitly or implicitly distributed. If they are explicitly distributed, their entries have a specific distribution. Otherwise, the entries have an implicit distribution imposed by generative algorithm the matrix uses.

## Explicitly Distributed

For (homogenous) explicitly distributed matrices, we can use a ``function factory'' method to be concise. The actual implementation is more verbose for the purposes of argument documentation, but the following code is minimal and fully functional. Additionally, there are the beta matrices, which use the matrix model provided by the algorithm in Dimitriu's paper.

### Homogenously Distributed

```{r}
# ... represents all the arguments taken in by the rdist function
RM_explicit <- function(rdist){
  function(N, ..., symm = FALSE){
    # Create an [N x N] matrix sampling the rows from rdist, passing ... to rdist
    P <- matrix(rdist(N^2, ...), nrow = N)
    # Make symmetric if prompted
    if(symm){P <- .makeHermitian(P)}
    # Return P
    P
  }
}

# A version where we add an imaginary component
RM_explicit_cplx <- function(rdist){
  RM_dist <- function(N, ..., symm = FALSE, cplx = FALSE, herm = FALSE){
    # Create an [N x N] matrix sampling the rows from rdist, passing ... to rdist
    P <- matrix(rdist(N^2, ...), nrow = N)
    # Make symmetric/hermitian if prompted
    if(symm || herm){P <- .makeHermitian(P)}
    # Returns a matrix with complex (and hermitian) entries if prompted
    if(cplx){
      # Recursively add imaginary components as 1i * instance of real-valued matrix.
      Im_P <- (1i * RM_dist(N, ...))
      # Make imaginary part Hermitian if prompted
      if(herm){P <- P + .makeHermitian(Im_P)}
      else{P <- P + Im_P}
    }
    P # Return the matrix
  }
}
```

With our function factories set up, we can quickly generate all the random matrix functions for all the distributions our hearts could desire.

```{r}
RM_unif <- RM_explicit_cplx(runif)
RM_norm <- RM_explicit_cplx(rnorm)
```


### Beta Matrices

For the $\beta$-ensemble matrices, we simply use the algorithm provided in Dimitriu's paper. Doing so, we get the function:

```{r}
# Generate a Hermite beta matrix using Dumitriu's Matrix Model
RM_beta <- function(N, beta){
  # Set the diagonal ~ N(0,2)
  P <- diag(rnorm(n = N, mean = 0, sd = sqrt(2)))
  # Get degrees of freedom sequence for offdigonal
  df_seq <- beta * (N - seq(1, N-1))
  # Set the off-1 diagonals as chi squared variables with df(beta_i)
  P[row(P) - col(P) == 1] <- P[row(P) - col(P) == -1] <- sqrt(rchisq(N-1, df_seq))
  # Rescale the entries by 1/sqrt(2)
  P <- P/sqrt(2)
  # Return the beta matrix
  P
}
```

\newpage

## Implicitly Distributed

In the case of implicitly distributed matrices, we have various types of stochastic matrices.

### Stochastic Matrices

For stochastic matrices, we require slightly more setup. First, we setup the row functions to sample probability vectors:

```{r}
# Generates stochastic rows of length N
.stoch_row <- function(N){
  # Sample a vector of probabilities
  row <- runif(n = N, min = 0, max = 1)
  # Return the normalized row (sums to one)
  row / sum(row)
}
```

For random introduced sparsity, we define the following row function.

```{r}
# Generates same rows as in .stoch_row(N), but with randomly introduced sparsity
.stoch_row_zeros <- function(N){
  # Sample a vector of probabilities
  row <- runif(n = N, min = 0, max = 1)
  # Sample a vertex degree of at least one (as to ensure row is stochastic)
  degree_vertex <- sample(x = 1:(N-1), size = 1)
  # Sever a random selection of edges to set the vertex degree
  row[sample(1:N, size = N - degree_vertex)] <- 0
  # Return normalized row
  row / sum(row)
}
```

Once this is done, we can use this function iteratively. With some magic, we can incorporate an option to make the matrix symmetric, and we get the following function.

```{r}
RM_stoch <- function(N, symm = F, sparsity = F){
  # Choose row function depending on sparsity argument
  if(sparsity){row_fxn <- .stoch_row_zeros} else {row_fxn <- .stoch_row}
  # Generate the [N x N] stochastic matrix stacking N stochastic rows
  P <- do.call("rbind", lapply(X = rep(N, N), FUN = row_fxn))
  # Make symmetric (if prompted)
  if(symm){ P <- .makeStochSymm(P) }
  # Return the matrix
  P
}
```

### Erdos-Renyi Matrices

For the Erdos-Renyi walks, we do something similar by defining a parameterized row function.

```{r}
# Generates a stochastic row with parameterized sparsity of p
.stoch_row_erdos <- function(N, p){
  # Sample a vector of probabilities
  row <- runif(n = N, min = 0, max = 1)
  # Sample the vertex degree so that it is ~ Bin(n,p)
  degree_vertex <- rbinom(n = 1, size = N, prob = 1 - p)
  # Sever a random selection of edges to set the vertex degree
  row[sample(1:N, degree_vertex)] <- 0
  # Return normalized row only if non-zero (cannot divide by 0)
  if(sum(row) != 0){
    row / sum(row)
  } else{
    .stoch_row_erdos(N, p) # Otherwise, try again
  }
}
```

And we again use the row function iteratively to get the following function.

```{r}
RM_erdos <- function(N, p, stoch = T){
  # Generate an [N x N] Erdos-Renyi stochastic matrix by stacking N p-stochastic rows
  P <- do.call("rbind", lapply(X = rep(N, N), FUN = .stoch_row_erdos, p = p))
  # Return the Erdos-Renyi transition matrix
  P
}
```

And as such, we have minimal, functional implementations of functions that sample random matrices! In total, we only needed two helper functions. The `.offdiagonalEntries` function was used to normalize the probabilities in `RM_stoch` and `RM_erdos`.

```{r}
# Returns a Hermitian version of a matrix by manual assignment
.makeHermitian <- function(P){
  for(i in 1:nrow(P)){
    for(j in 1:ncol(P)){
      # Select the entries in the upper triangle (i < j)
      if(i < j){
        # Make the upper triangle equal to the conjugate transpose of the lower triangle
        P[i,j] <- Conj(P[j,i])
      }
    }
  }
  # Return the Hermitian Matrix
  P
}

# Return the off-diagonal entries of row i
.offdiagonalEntries <- function(row, row_index){row[which(1:length(row) != row_index)]}
```

\newpage

## Ensemble Extensions

Lastly, we have the ensemble extensions. These functions are quite simple to implement user a ``function factory''. Again, the actual implementations are more verbose due to the argument descriptions, but otherwise, are exactly the same.

```{r}
RME_extender <- function(RM_dist){
  # Function returns a list of replicates of the RM_dist function with ... as its arguments
  function(N, ..., size){
    lapply(X = rep(N, size), FUN = RM_dist, ...)
  }
}
```

Now, we extend the functions as follows, and we are done with the matrix module!

```{r}
RME_unif <- RME_extender(RM_unif)
RME_norm <- RME_extender(RM_norm)
RME_beta <- RME_extender(RM_beta)
RME_stoch <- RME_extender(RM_stoch)
RME_erdos <- RME_extender(RM_erdos)
```

\newpage

# Spectral Statistics

## Spectrum

```{r}
spectrum <- function(array, components = T, norm_order = T, singular = F, order = NA){
  # Digits to round values to
  digits <- 4
  # Get the type of array
  array_class <- .arrayClass(array)
  # For ensembles, iteratively rbind() each matrix's spectrum
  if(array_class == "ensemble"){
    purrr::map_dfr(array, .spectrum_matrix, components, norm_order, singular, order, digits)
  }
  # From matrices, call the function returning the ordered spectrum for a singleton matrix
  else if(array_class == "matrix"){
    .spectrum_matrix(array, components, norm_order, singular, order, digits)
  }
}
```

```{r}
# Helper function returning tidied eigenvalue array for a matrix
.spectrum_matrix <- function(P, components, norm_order, singular, order, digits = 4){
  # For singular values, take P as product of the itself and its tranpose
  if(singular){P <- P %*% t(P)}
  # Get the eigenvalues of P
  eigenvalues <- eigen(P, only.values = TRUE)$values
  # Take the square root of the eigenvalues to obtain singular values
  if(singular){eigenvalues <- sqrt(eigenvalues)}
  # Sort the eigenvalues to make it an ordered spectrum
  eigenvalues <- .sortValues(eigenvalues, norm_order)
  # If uninitialized, select all orders; otherwise, use c() so singletons => vectors
  if(class(order) == "logical"){order <- 1:nrow(P)} else{order <- c(order)}
  # Return the spectrum of the matrix
  return(purrr::map_dfr(order, .resolve_eigenvalue, eigenvalues, components, digits))
}
```

```{r}
# Read and parse an eigenvalue from a sorted eigenvalue array
.resolve_eigenvalue <- function(order, eigenvalues, components, digits){
  # Read from a sorted eigenvalue array at that order
  eigenvalue <- eigenvalues[order]
  # Get norm and order columns
  features <- data.frame(Norm = abs(eigenvalue), Order = order)
  if(components){
    # If components are sought, resolve the eigenvalue into seperate columns first
    res <- cbind(data.frame(Re = Re(eigenvalue), Im = Im(eigenvalue)), features)
  } else{
    # Otherwise, don't resolve the eigenvalue components
    res <- cbind(data.frame(Eigenvalue = eigenvalue), features)
  }
  # Round entries and return the resolved eigenvalue
  res <- round(res, digits)
  return(res)
}
```

### Helper Functions

```{r}
# Parses an array to see classify it as a matrix or an ensemble of matrices.
.arrayClass <- function(array){
  # Sample an element from the array and get its class
  elem <- array[[1]]
  types <- class(elem)
  # Classify it by analyzing the element class
  if("numeric" %in% types || "complex" %in% types){
    return("matrix")
  }
  else if("matrix" %in% types){
    return("ensemble")
  }
}

# Sort an array of numbers by their norm (written for eigenvalue sorting)
.sortValues <- function(vals, norm_order){
  values <- data.frame(value = vals)
  # If asked to sort by norms, arrange by norm and return
  if(norm_order){
    values$norm <- abs(values$value)
    values <- values %>% arrange(desc(norm))
    # Return the norm-sorted values
    values$value
  }
  # Otherwise, sort by sign and return
  else{ sort(vals, decreasing = TRUE) }
}
```


\newpage

## Dispersions

```{r}
# Compute the dispersion of a matrix or matrix ensemble
dispersion <- function(array, pairs = NA, norm_order = T, singular = F, pow_norm = 1){
  # Digits to round values to
  digits <- 4
  # Get the type of array
  array_class <- .arrayClass(array)
  # Parse input and generate pair scheme (default NA), passing on array for dimension
  pairs <- .parsePairs(pairs, array, array_class)
  # For ensembles; iteratively rbind() each matrix's dispersion
  if(array_class == "ensemble"){
    purrr::map_dfr(array, .dispersion_matrix, pairs, norm_order, singular, pow_norm, digits)
  }
  # Array is a matrix; call function returning dispersion for singleton matrix
  else if(array_class == "matrix"){
    .dispersion_matrix(array, pairs, norm_order, singular, pow_norm, digits)
  }
}
```

```{r}
# Find the eigenvalue dispersions for a given matrix
.dispersion_matrix <- function(P, pairs, norm_order, singular, pow_norm, digits = 4){
  # Get the ordered spectrum of the matrix
  eigenvalues <- spectrum(P, norm_order = norm_order, singular = singular)
  # Generate norm function to pass along as argument (Euclidean or Beta norm)
  norm_fn <- function(x){ (abs(x))^pow_norm }
  # Compute and return the dispersion
  purrr::map2_dfr(pairs[["i"]], pairs[["j"]], .resolve_dispersion, eigenvalues, norm_fn, digits)
}
```

```{r}
# Read and parse a dispersion observation between eigenvalue i and j.
.resolve_dispersion <- function(i, j, eigenvalues, norm_fn, digits){
  # Initialize dispersion dataframe by adding order of eigenvalues compared
  disp <- data.frame(i = i, j = j)
  # Add the eigenvalues
  disp$eig_i <- .read_eigenvalue(i, eigenvalues)
  disp$eig_j <- .read_eigenvalue(j, eigenvalues)
  # Get the identity difference
  disp$id_diff <- disp$eig_j - disp$eig_i
  # Compute norm of the identity difference (standard norm metric)
  disp$id_diff_norm <- norm_fn(disp$id_diff)
  # Compute the difference of absolutes
  disp$abs_diff <- norm_fn(disp$eig_j) - norm_fn(disp$eig_i)
  # Round digits
  disp <- round(disp, digits)
  # Get the ranking difference
  disp$diff_ij <- disp$i - disp$j
  # Return the resolved dispersion observation
  disp
}
```

\newpage

### Helper Functions

```{r}
# Parse a string argument for which pairing scheme to utilize
.parsePairs <- function(pairs, array, array_class){
  # Valid schemes for printing if user is unaware of options
  valid_schemes <- c("largest", "lower", "upper", "consecutive", "all")
  # Set default to be the consecutive pair scheme
  if(class(pairs) == "logical"){pairs <- "consecutive"}
  # Stop function call if the argument is invalid
  if(!(pairs %in% valid_schemes)){
    scheme_list <- paste(valid_schemes, collapse = ", ")
    stop(paste("Invalid pair scheme. Try one of the following: ", scheme_list, ".", ""))
  }
  # // Once we verify that we have a valid pair scheme string, try to parse it.
  # First, obtain a matrix by inferring array type; if ensemble take first matrix
  if(array_class == "ensemble") { P <- array[[1]] }
  else if(array_class == "matrix") { P <- array }
  # Obtain the dimension of the matrix
  N <- nrow(P)
  # Parse the pair string and evaluate the pair scheme
  if(pairs == "largest"){pair_scheme <- data.frame(i = 2, j = 1)}
  else if(pairs == "consecutive"){pair_scheme <- .consecutive_pairs(N)}
  else if(pairs == "lower"){pair_scheme <- .unique_pairs_lower(N)}
  else if(pairs == "upper"){pair_scheme <- .unique_pairs_upper(N)}
  else if(pairs == "all"){pair_scheme <- .all_pairs(N)}
  # Return pair scheme
  return(pair_scheme)
}
```

\newpage

### Pairing Schema

```{r}
# The trivial pairing scheme:
# Enumerate all possible pairs.
.all_pairs <- function(N){
  purrr::map_dfr(1:N, function(i, N){data.frame(i = rep(i, N), j = 1:N)}, N)
}
#=================================================================================#
# The consecutive pairing scheme:
# Enumerate all possible consecutive/neighboring pairs. Ensures no linear combiantions.
.consecutive_pairs <- function(N){
  purrr::map_dfr(2:N, function(i){data.frame(i = i, j = as.integer(i - 1))})
}
#=================================================================================#
# The lower-triangular pairing scheme:
# Enumerate the pair combinations given N items with i > j.
.unique_pairs_lower <- function(N){
  is <- do.call("c", purrr::map(1:N, function(i){rep(i,N)}))
  js <- rep(1:N, N)
  # Helper function: selects elements only if they are lower triangular
  .isLowerTri <- function(i, j){if(i > j){ c(i = i, j = j) }}
  pairs <- do.call("rbind",purrr::map2(is, js, .f = .isLowerTri))
  data.frame(pairs)
}
#=================================================================================#
# The upper-triangular pairing scheme:
# Enumerate the pair combinations given N items with i < j.
.unique_pairs_upper <- function(N){
  is <- do.call("c", purrr::map(1:N, function(i){rep(i,N)}))
  js <- rep(1:N, N)
  # Helper function: selects elements only if they are lower triangular
  .isUpperTri <- function(i, j){if(i < j){ c(i = i, j = j) }}
  pairs <- do.call("rbind",purrr::map2(is, js, .f = .isUpperTri))
  data.frame(pairs)
}
```

\newpage

## Parallel Extensions

### Spectrum

```{r}
spectrum_parallel <- function(array, components = TRUE, sort_norms = TRUE, singular = FALSE, order = NA){
  digits <- 4 # Digits to round values to
  # Array is a matrix; call function returning eigenvalues for singleton matrix
  if(class(array) == "matrix"){
    .spectrum_matrix(array, components, sort_norms, singular, order, digits)
  }
  # Array is an ensemble; recursively row binding each matrix's eigenvalues
  else if(class(array) == "list"){
    furrr::future_map_dfr(array, .spectrum_matrix, components, sort_norms, singular, order, digits)
  }
}
```

### Dispersion

```{r}
dispersion_parallel <- function(array, pairs = NA, sort_norms = TRUE, singular = FALSE, norm_pow = 1){ #sortNorms? orderByNorms? pair_scheme?
  digits <- 4 # Digits to round values to
  pairs <- .parsePairs(pairs, array) # Parse input and generate pair scheme (default NA), passing on array for dimension and array type inference
  # Array is a matrix; call function returning dispersion for singleton matrix
  if(class(array) == "matrix"){
    .dispersion_matrix(array, pairs, sort_norms, singular, norm_pow, digits)
  }
  # Array is an ensemble; recursively row binding each matrix's dispersions
  else if(class(array) == "list"){
    furrr::future_map_dfr(array, .dispersion_matrix, pairs, sort_norms, singular, norm_pow, digits)
  }
}
```
